---
title: "Chapter 3 Exercises"
author: "Darrell A. Carvalho"
output:
  html_document: 
    toc: yes
    toc_depth: 4
    toc_float: yes
    theme: spacelab
    number_sections: yes
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, eval = TRUE, include = TRUE}
knitr::opts_chunk$set(echo = TRUE)
# install.packages("remotes")
# remotes::install_github("geocompr/geocompkg")
pacman::p_load(sf, terra, spData, dplyr,tmap)
data(us_states)
data(us_states_df)
```

# Exercise 1:

Create a new object called `us_states_name` that contains only the
`NAME` column from the `us_states` object using either base R (`[`) or
tidyverse(`select()`) syntax. What is the cost of the new object and
what makes it geographic?

```{r}
us_states_name <- us_states %>% select(NAME)
class(us_states_name)
# the sf column is sticky, and is what keeps data.frames spatial
```

# Exercise 2:

Select columns from the `us_states` object which contain population
data. Obtain the same result using a different command (bonus: try to
find three ways of obtaining the same result). Hint: try to use helper
functions, such as `contains` or `matches` from **dplyr** (see
`?contains`).

```{r}
us_states %>% select(contains("Pop"))
us_states %>% select(starts_with("total_pop"))
us_states[,5:6]
us_states %>% select(matches("total"))
```

# Exercise 3:

Find all states with the following characteristics (bonus find and plot
them):

-   Belong to the Midwest region.

-   Belong to the West region, have an area below 250,000 km2and in 2015
    a population greater than 5,000,000 residents (hint: you may need to
    use the function units::set_units() or as.numeric()).

-   Belong to the South region, had an area larger than 150,000 km2 or a
    total population in 2015 larger than 7,000,000 residents.

```{r}
# Midwest States
us_states %>%
  filter(REGION=="Midwest") %>%
  qtm(text="NAME")
# Western States with an area under 250K km^2 and a 2015 population over 5 mil
us_states %>%
  filter(REGION=="West" & (as.numeric(AREA) < 250000 & total_pop_15 > 5000000)) %>%
  qtm(text="NAME", fill="total_pop_15")
# Southern states with an area over 150K km2 and a 2015 pop over 7 mil
us_states %>%
  filter(REGION=="South" & as.numeric(AREA) > 150000 & total_pop_15 > 7000000) %>%
  qtm(text="NAME", fill="total_pop_15")

```

# Exercise 4:

What was the total population in 2015 in the `us_states` dataset? What
was the minimum and maximum total population in 2015?

```{r}
us_states$total_pop_15 %>% sum()
us_states %>% 
  st_drop_geometry() %>% 
  select(total_pop_15) %>% 
  summary()
```

# Exercise 5:

How many states are there in each region?

```{r}
us_states %>% 
  group_by(REGION) %>% 
  count()
```

# Exercise 6:

What was the minimum and maximum total population in 2015 in each
region? What was the total population in 2015 in each region?

```{r}
us_states %>% 
  st_drop_geometry() %>% 
  group_by(REGION) %>% 
  summarise(min = min(total_pop_15), 
            max = max(total_pop_15), 
            total = sum(total_pop_15))
```

# Exercise 7:

Add variables from us_states_df to us_states, and create a new object
called us_states_stats. What function did you use and why? Which
variable is the key in both datasets? What is the class of the new
object?

```{r}
us_state_stats <- us_states %>% 
  left_join(us_states_df, by = c("NAME" = "state"))
class(us_state_stats) # sticky sf
```

# Exercise 8:

`us_states_df` has two more rows than `us_states`. How can you find them? (hint: try to use the `dplyr::anti_join()` function)

```{r}
us_states_df %>% anti_join(us_states, by = c("state" = "NAME"))
```

# Exercise 9:

What was the population density in 2015 in each state? What was the population density in 2010 in each state?

```{r}
us_states_dens <- us_states %>% mutate(pop_dens_10 = total_pop_10/as.numeric(AREA), pop_dens_15 = total_pop_15/as.numeric(AREA))
```

# Exercise 10:

How much has population density changed between 2010 and 2015 in each state? Calculate the change in percentages and map them.

```{r}
us_states_dens %>%
  mutate(pop_dens_change = ((pop_dens_15-pop_dens_10)/pop_dens_10) * 100) %>%
  qtm(fill = "pop_dens_change")
```
# Exercise 11:
Change the columns’ names in `us_states` to lowercase. (Hint: helper functions - `tolower()` and `colnames()` may help.)

```{r}
us_states %>% 
  colnames %>% 
  tolower()
```

