---
title: "Chapter 3 Exercises"
author: "Darrell A. Carvalho"
output:
  html_document: 
    toc: yes
    toc_depth: 4
    toc_float: yes
    theme: spacelab
    number_sections: yes
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, eval = TRUE, include = TRUE}
knitr::opts_chunk$set(echo = TRUE)
# install.packages("remotes")
# remotes::install_github("geocompx/geocompkg")
pacman::p_load(sf, terra, spData, dplyr,tmap)
data(us_states)
data(us_states_df)
```

# Exercise 1:

Create a new object called `us_states_name` that contains only the
`NAME` column from the `us_states` object using either base R (`[`) or
tidyverse(`select()`) syntax. What is the cost of the new object and
what makes it geographic?

```{r}
us_states_name <- us_states %>% select(NAME)
class(us_states_name)
# the sf column is sticky, and is what keeps data.frames spatial
```

# Exercise 2:

Select columns from the `us_states` object which contain population
data. Obtain the same result using a different command (bonus: try to
find three ways of obtaining the same result). Hint: try to use helper
functions, such as `contains` or `matches` from **dplyr** (see
`?contains`).

```{r}
us_states %>% select(contains("Pop"))
us_states %>% select(starts_with("total_pop"))
us_states[,5:6]
us_states %>% select(matches("total"))
```

# Exercise 3:

Find all states with the following characteristics (bonus find and plot
them):

-   Belong to the Midwest region.

-   Belong to the West region, have an area below 250,000 km2and in 2015
    a population greater than 5,000,000 residents (hint: you may need to
    use the function units::set_units() or as.numeric()).

-   Belong to the South region, had an area larger than 150,000 km2 or a
    total population in 2015 larger than 7,000,000 residents.

```{r}
# Midwest States
us_states %>%
  filter(REGION=="Midwest") %>%
  qtm(text="NAME")
# Western States with an area under 250K km^2 and a 2015 population over 5 mil
us_states %>%
  filter(REGION=="West" & (as.numeric(AREA) < 250000 & total_pop_15 > 5000000)) %>%
  qtm(text="NAME", fill="total_pop_15")
# Southern states with an area over 150K km2 and a 2015 pop over 7 mil
us_states %>%
  filter(REGION=="South" & as.numeric(AREA) > 150000 & total_pop_15 > 7000000) %>%
  qtm(text="NAME", fill="total_pop_15")

```

# Exercise 4:

What was the total population in 2015 in the `us_states` dataset? What
was the minimum and maximum total population in 2015?

```{r}
us_states$total_pop_15 %>% sum()
us_states %>% 
  st_drop_geometry() %>% 
  select(total_pop_15) %>% 
  summary()
```

# Exercise 5:

How many states are there in each region?

```{r}
us_states %>% 
  group_by(REGION) %>% 
  count()
```

# Exercise 6:

What was the minimum and maximum total population in 2015 in each
region? What was the total population in 2015 in each region?

```{r}
us_states %>% 
  st_drop_geometry() %>% 
  group_by(REGION) %>% 
  summarise(min = min(total_pop_15), 
            max = max(total_pop_15), 
            total = sum(total_pop_15))
```

# Exercise 7:

Add variables from `us_states_df` to `us_states`, and create a new object
called `us_states_stats`. What function did you use and why? Which
variable is the key in both datasets? What is the class of the new
object?

```{r}
us_state_stats <- us_states %>% 
  left_join(us_states_df, by = c("NAME" = "state"))
class(us_state_stats) # sticky sf
```

# Exercise 8:

`us_states_df` has two more rows than `us_states`. How can you find them? (hint: try to use the `dplyr::anti_join()` function)

```{r}
us_states_df %>% anti_join(us_states, by = c("state" = "NAME"))
```

# Exercise 9:

What was the population density in 2015 in each state? What was the population density in 2010 in each state?

```{r}
us_states_dens <- us_states %>% mutate(pop_dens_10 = total_pop_10/as.numeric(AREA), pop_dens_15 = total_pop_15/as.numeric(AREA))
us_states_dens
```

# Exercise 10:

How much has population density changed between 2010 and 2015 in each state? Calculate the change in percentages and map them.

```{r}
us_states_dens %>%
  mutate(pop_dens_change = ((pop_dens_15-pop_dens_10)/pop_dens_10) * 100) %>%
  qtm(fill = "pop_dens_change")
```

# Exercise 11:

Change the columns’ names in `us_states` to lowercase. (Hint: helper functions - `tolower()` and `colnames()` may help.)

```{r}
us_states %>% 
  colnames %>% 
  tolower()
```
# Exercise 12:
Using `us_states` and `us_states_df` create a new object called `us_states_sel`. The new object should have only two variables - `median_income_15` and `geometry`. Change the name of the `median_income_15` column to Income.

```{r}
us_states_sel <- us_states %>% 
  left_join(us_states_df, by = c("NAME" = "state")) %>% 
  transmute(Income = median_income_15)
us_states_sel
```

# Exercise 13:

Calculate the change in the number of residents living below the poverty level between 2010 and 2015 for each state. (Hint: See `?us_states_df` for documentation on the poverty level columns.) Bonus: Calculate the change in the percentage of residents living below the poverty level in each state.

```{r}
us_pov_pct <- us_state_stats %>% mutate(pct_blw_pvt_10 =
                               poverty_level_10 / total_pop_10 * 100,
                             pct_blw_pvt_15 = 
                               poverty_level_15 / total_pop_15 * 100,
                             pct_pt_change = (pct_blw_pvt_15 - pct_blw_pvt_10))
us_pov_pct
```

# Exercise 14:

What was the minimum, average and maximum state’s number of people living below the poverty line in 2015 for each region? Bonus: What is the region with the largest increase in people living below the poverty line?

```{r region-poverty-summary}
us_state_stats %>%
  group_by(REGION) %>% 
  summarise(min = min(poverty_level_15), 
            max = max(poverty_level_15), 
            average = mean(poverty_level_15))
```
```{r region-poverty-change}
us_state_stats %>%
  group_by(REGION) %>% 
  summarise(region_pov_rate_10 =
              sum(poverty_level_10) /sum(total_pop_10) * 100,
            region_pov_rate_15 =
              sum(poverty_level_15) /sum(total_pop_15) * 100,
            pct_pt_change = region_pov_rate_15 - region_pov_rate_10
            ) %>% st_drop_geometry()
```

# Exercise 15:

Create a raster from scratch with nine rows and columns and a resolution of 0.5 decimal degrees (WGS84). Fill it with random numbers. Extract the values of the four corner cells.
```{r}
rasta <- rast(nrows = 9,ncols = 9,crs = "epsg:4326",xmin = 0,xmax = 4.5,ymin = 0,ymax = 4.5,vals = rnorm(81))
c(rasta[1,1], rasta[1,9], rasta[9,9], rasta[9, 1])
```

# Exercise 16:

What is the most common class of our example raster `grain`?

```{r}
grain_order = c("clay", "silt", "sand")
grain_char = sample(grain_order, 36, replace = TRUE)
grain_fact = factor(grain_char, levels = grain_order)
grain = rast(nrows = 6, ncols = 6, 
             xmin = -1.5, xmax = 1.5, ymin = -1.5, ymax = 1.5,
             vals = grain_fact)
cats(grain)
```


# Exercise 17:

Plot the histogram and the boxplot of the `dem.tif` file from the spDataLarge package (`system.file("raster/dem.tif", package = "spDataLarge")`).

```{r}
dem <- rast(system.file("raster/dem.tif", package = "spDataLarge"))
dem %>% hist()
dem %>% boxplot()
```

