---
title: "Chapter 3 Notes"
author: "Darrell A. Carvalho"
output:
  html_document: 
    toc: yes
    toc_depth: 4
    toc_float: yes
    theme: spacelab
    number_sections: yes
---

# Setup

The code in the following chunk installs all the packages necessary for reproducing the book.

The authors demonstrate loading libraries using the `library()` function, but I prefer using pacman's `p_load()`

```{r setup, eval = TRUE, include = TRUE}
knitr::opts_chunk$set(echo = TRUE)
# install.packages("remotes")
# remotes::install_github("geocompr/geocompkg")
pacman::p_load(sf, terra, spData, dplyr)
```

------------------------------------------------------------------------

# Introduction

-   Attribute data - non-spatial information associated with geometry data
    -   in raster data, this is usually stored in the header
    -   in vector data, this is the tabular data outside of the geometry columns

# Vector attribute manipulation

-   `sf` class extends base `data.frame`
    -   one column per variable
    -   one row per observation/*feature*
    -   `geometry` column
        -   `sfc` class
-   `sf` class provides generic methods to work like regular data frames

```{r generic-methods}
methods(class = "sf")
```

-   `sf` objects store spatial and non-spatial data as columns in `data.frame`
-   `sf` objects can extend `tibble`

```{r world-base-functions}
class(world) # sf object and tibble
dim(world) # 2 dimensions, 177 rows, 11 columns
```

-   drop geometry to turn `sf` object into `tibble`

    -   useful for speeding up non-spatial data manipulation

    ```{r drop-geometries}
    world_df <- st_drop_geometry(world)
    class(world_df) # tibble data frame
    ncol(world_df) # 10 columns (no geometry)
    ```

    ## Vector attribute subsetting

-   subset with brackets `[]` or base function`subset()`

    -   `[]` subsets rows or columns
        -   `world[i, j]` returns rows *i* and columns *j*

            -   leaving one or other blank returns all rows or all columns

            ```{r base-subsetting, echo=TRUE, message=TRUE, paged.print=TRUE}
            world[1:6, ] # rows 1:6, all columns
            world[, 1:3] # columns 1:3, all rows, + geometry column because 'sticky' sf
            world[1:6, 1:3] # rows 1:6, columns 1:3 + geometry
            world[,c("name_long", "pop")] # return columns by name, + geometry
            world[, c(T,T,F,F,F,F,F,T,T,F,F)] # logic-based subsetting of columns
            # world[, 888] throws an error because the column doesn't exist
            ```

**The following code demonstrates usefulness of logical subsetting**

```{r logic-subset, echo=FALSE}
small_countries <- world[world$area_km2 < 10000,]
# same as
small_countries <- subset(world, area_km2 < 10000)
```

-   `dplyr` subsetting can be done with:

    -   `dplyr::filter()` or `dplyr::slice()` for rows
    -   `dplyr::select()` for columns
    -   both are non-destructive to spatial components

    ```{r dplyr-select}
    world1 <- dplyr::select(world, name_long, pop) # select name_long and pop cols
    names(world1)

    world2 <- dplyr::select(world, name_long:pop) # all columns between name_long and pop (inclusive)
    names(world2)

    world3 <- dplyr::select(world, -subregion, -area_km2) # all columns except subregion and area_km2 (inclusive)
    names(world3)

    world4 <- dplyr::select(world, name_long, population = pop) # renaming a column through selection
    names(world4)

    world5 <- world[, c("name_long", "pop")] # subset columns by name
    names(world5)[names(world5) == "pop"] = "population" # rename column manually via base R
    names(world5)
    ```

-   columns can be returned as vectors via `pull()` and `dataset$column` or `dataset[["column"]]` syntaxes

```{r vector-return}
pull(world, pop)
world$pop
world[["pop"]]
```

```{r using-slice}
slice(world, 1:6) # slice operates on rows, similar to how select operates on columns
```

## Chaining commands with pipes

-   pipe allows for chaining
-   system pipe is `|>`; tidy pipe is `%>%`

```{r chaining}
world7 <- world %>%
  filter(continent == "Asia") %>%
  dplyr::select(name_long, continent) %>%
  slice(1:5)
```

-   piping is much more readable in many peoples' minds than nesting (see below)

```{r nesting}
world8 <- slice(
  dplyr::select(
    filter(world, continent == "Asia"),
    name_long, continent),
  1:5)
```

-   splitting operations into self-contained lines is useful for debugging or package development.

```{r split-lines}
world9_filtered <- filter(world, continent == "Asia")
world9_selected <- dplyr::select(world9_filtered, continent)
world9 <- slice(world9_selected, 1:5)
```

## Vector attribute aggregation

-   Aggregation summarizes data into grouping variables
    -   Calculate continent population based on country-level data
-   base aggregate() will return either a non-spatial data.frame or a sf object, depending on the order used in the function call

```{r cont-pop-nonspatial}
world_agg1 <- aggregate(pop ~ continent, FUN = sum, data = world, na.rm = TRUE)
class(world_agg1)
```

```{r cont-pop-spatial}
world_agg2 <- aggregate(world["pop"], list(world$continent), FUN = sum,
                        na.rm = TRUE)
class(world_agg2)
```

-   dplyr aggregation uses group_by()

```{r dplyr-aggregation}
world %>% 
  group_by(continent) %>% 
  summarize(pop = sum(pop, na.rm = TRUE)) -> 
  world_agg3
class(world_agg3) # outputs a tibble
```

-   this approach allows more flexability in output generation; see below:

```{r dplyr-aggregation-computation}
world %>% 
  group_by(continent) %>% 
  summarize(
    pop = sum(pop, na.rm = TRUE), # sums the population to the continent
    `area_sqkm` = sum(area_km2), # finds the area of the continent
    n = n()) -> # counts the number of countries in each continent
  world_agg4
world_agg4
```

```{r aggregation-chaining}
world %>% 
  st_drop_geometry() %>% # remove spatial column(s)
  dplyr::select(pop, continent, area_km2) %>%  # keep columns we care about
  group_by(continent) %>% # Aggregation value
  summarize(Pop = sum(pop, na.rm = TRUE), # summarize pop
            Area = sum(area_km2),         # summarize area
            N = n()) %>%                  # count of countries
  mutate(Density = round(Pop / Area)) %>% # compute pop density
  slice_max(Pop, n = 3) %>%               # keep top three
  arrange(desc(N)) ->                     # sort by number of countries
  world_agg5
world_agg5
```

## Vector attribute joining

-   utilizes common table joins from relational data models
    -   dplyr provides `left_join() and`inner_join()\` among others
    -   these joins follow SQL conventions

```{r coffee-join}
world_coffee <- left_join(world, coffee_data) # left_join preserves first set
class(world_coffee)

names(world_coffee)

plot(world_coffee["coffee_production_2017"])
```

-   Joins require a key variable in both/all datasets
    -   dplyr uses any var with name matches
    -   where mismatches occur, either rename the key variable in one set or use by argument to specify join var

```{r join-with-by}
coffee_renamed <- rename(coffee_data, nm = name_long) # rename for demo
world_coffee2 <- left_join(world, coffee_renamed, by = c(name_long = "nm"))
```

-   With left_join, all rows from first set are kept; empty cells filled with N/A

-   inner_join keeps only values from original set that have a match with new set

```{r inner-join}
world_coffee_inner <- inner_join(world, coffee_data)
nrow(world_coffee_inner)
setdiff(coffee_data$name_long, world$name_long)
# DRC is missing because of abbreviation
```

```{r data-preprocessing-for-fixed-join}
drc <- stringr::str_subset(world$name_long, "Dem*.+Congo") # correct name
coffee_data$name_long[grepl("Congo,", coffee_data$name_long)] = drc # fix coffee_data
```

```{r inner-join-redux}
world_coffee_match <- inner_join(world, coffee_data)
nrow(world_coffee_match)
```

-   operating in the other direction will return a tibble, as output tends to match the first input in the function

```{r joining-position-swapped}
coffee_world <- left_join(coffee_data, world) # coffee first, world second
class(coffee_world)

# Coerce it back to spatial with st_as_sf()
st_as_sf(coffee_world)
```

## Creating attributes and removing spatial information
- base R attribute creation
```{r base-r-area-column}
world_new <- world # prevent original data corruption
world_new$pop_dens <- world_new$pop / world_new$area_km2
```

- dplyr provides `mutate()` and `transmute()` functions
  - `mutate()` adds new columns at the end
  - `transmute()` drops non-specified columns
```{r using-mutate}
world %>% 
  mutate(pop_dens = pop / area_km2)
```

```{r using-transmute}
world %>% 
  transmute(pop_dens = pop / area_km2)
```

- `tidyr::unite()` joins existing columns
  - `sep` argument defines a separator
  - `remove` argument determines whether to drop original columns

```{r tidyr-unite}
world %>% 
  tidyr::unite("con_reg",
               continent:region_un,
               sep = ":",
               remove = TRUE) ->
  world_unite
```

