---
title: "Chapter 3 Notes"
author: "Darrell A. Carvalho"
output:
  html_document: 
    toc: yes
    toc_depth: 4
    toc_float: yes
    theme: spacelab
    number_sections: yes
---
# Setup

The code in the following chunk installs all the packages necessary for reproducing the book.

The authors demontrate loading libraries using the `library()` function, but I prefer using pacman's `p_load()`

```{r setup, eval = TRUE, include = TRUE}
knitr::opts_chunk$set(echo = TRUE)
# install.packages("remotes")
# remotes::install_github("geocompr/geocompkg")
pacman::p_load(sf, terra, spData, dplyr)
```

------------------------------------------------------------------------
# Introduction
-   Attribute data - non-spatial information associated with geometry data
    - in raster data, this is usually stored in the header
    - in vector data, this is the tabular data outside of the geometry columns

# Vector attribute manipulation
-   `sf` class extends base `data.frame`
    +   one column per variable
    +   one row per observation/*feature*
    +   `geometry` column
        *   `sfc` class
-   `sf` class provides generic methods to work like regular data frames
```{r generic-methods}
methods(class = "sf")
```
-   `sf` objects store spatial and non-spatial data as columns in `data.frame`
-   `sf` objects can extend `tibble`
```{r world-base-functions}
class(world) # sf object and tibble
dim(world) # 2 dimensions, 177 rows, 11 columns
```
-   drop geometry to turn `sf` object into `tibble`
    - useful for speeding up non-spatial data manipulation
```{r drop-geometries}
world_df <- st_drop_geometry(world)
class(world_df) # tibble data frame
ncol(world_df) # 10 columns (no geometry)
```
## Vector attribute subsetting
-   subset with brackets `[]` or base function`subset()`
    -   `[]` subsets rows or columns
        - `world[i, j]` returns rows *i* and columns *j*
            -   leaving one or other blank returns all rows or all columns
```{r base-subsetting, echo=TRUE, message=TRUE, paged.print=TRUE}
world[1:6, ] # rows 1:6, all columns
world[, 1:3] # columns 1:3, all rows, + geometry column because 'sticky' sf
world[1:6, 1:3] # rows 1:6, columns 1:3 + geometry
world[,c("name_long", "pop")] # return columns by name, + geometry
world[, c(T,T,F,F,F,F,F,T,T,F,F)] # logic-based subsetting of columns
# world[, 888] throws an error because the column doesn't exist
```


**The following code demonstrates usefulness of logical subsetting**
```{r logic-subset, echo=FALSE}
small_countries <- world[world$area_km2 < 10000,]
# same as
small_countries <- subset(world, area_km2 < 10000)
```

-   `dplyr` subsetting can be done with:
    -   `dplyr::filter()` or `dplyr::slice()` for rows
    -   `dplyr::select()` for columns
    - both are non-destructive to spatial components
```{r dplyr-select}
world1 <- dplyr::select(world, name_long, pop) # select name_long and pop cols
names(world1)
```

## Chaining commands with pipes
## Vector attribute aggregation
## Vector attribute joining
## Creating attributes and removing spatial information
