---
title: "Ch. 2 Exercises"
author: "Darrell A. Carvalho"
date: "3/14/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(spData)
library(sf)
library(terra)
library(spDataLarge)
```

# Ch. 2 Exercises
## E1: Use `summary()` on the geometry column of the `world` data obect that is included in the **spData** package. What does the output tell us about:
```{r E1, echo = FALSE}
summary(world$geom)
```
  * Its geometry type?
      + Multipolygon
  * The number of countries?
      + 177
  * Its coordinate reference system (CRS)?
      + WGS84 (epsg 4326)  

## E2: Run the code that 'generated' the map of the world in Section 2.2.3 (Basic map making).
```{r E2, echo=FALSE, warning=FALSE}
plot(world["continent"], reset = FALSE)
cex <- sqrt(world$pop) / 10000
world_cents  <- st_centroid(world, of_largest = TRUE)
plot(st_geometry(world_cents), add = TRUE, cex = cex)
```

  *  Find two similarities and two differences between the image on your computer and that in the book.
      - similarities: symbols scale with pop in both; both show increased distortion of shape nearer the poles.
      - differences: the book figure includes curved meridians; the population circles are centered in their respective countries' polygons in the book figure.
  * What does the `cex` argument do?
      - `cex` controls **c**haracter **ex**pansion - in this case by defining character expansion in proportion to a population factor.
  * Why was `cex` set to `sqrt(world$pop) / 10000`?
      - this creates a symbol scaling factor that scales in proportion to the population size.

## E3: Use `plot()` to create maps of Nigeria in context.
  * Adjust `lwd`, `col`, and `expandBB` arguments of `plot()`.
  * Challenge: read the documentation of `text()` and annotate the map.
```{r E3, echo=FALSE, warning=FALSE}
world_africa <- world[world$continent == "Africa",]
africa <- st_union(world_africa)
nigeria <- world[world$name_long == "Nigeria",] # subsets to only Nigeria
nigeria |>
  st_geometry() |>
  plot(main = "Nigeria in Context",
       expandBB = c(0,1.6,1,0), # expand BB at left and top
       col = "#008753", # Nigerian Green
       lwd = 4)
plot(world_africa[0], add = TRUE) # plots Africa underneath

name <- world[world$name_long == "Nigeria", ] # gets name data
namecenter <- st_centroid(name) # assigns the polygon centroid to the name
namecenter_coord <- st_coordinates(namecenter) # gets the centroid coordinates
text(x = namecenter_coord[, 1],
     y = namecenter_coord[, 2],
     labels = name$name_long,
     cex = 1.2) # plots the name annotation on the map
```

## E4: Create an empty `SpatRaster` object called `my_raster` with 10 columns and 10 rows. Assign random values between 0 and 10 to the new raster and plot it.
```{r E4, echo=FALSE, warning=FALSE}
my_raster <- rast(nrows = 10,
                  ncols = 10,
                  vals = sample(1:10, replace = TRUE, size = 100),
                  xmin = 0,
                  xmax = 10,
                  ymin = 0,
                  ymax = 10)
plot(my_raster)
```

## E5: Read-in the `raster/nlcd.tif` file from the **spDataLarge** package. What kind of information can you get about the properties of this file?
```{r E5, echo=FALSE, warning=FALSE}
nlcd_filepath <- system.file("raster/nlcd.tif", package = "spDataLarge")
nlcd_rast <- rast(nlcd_filepath)
nlcd_rast
plot(nlcd_rast)
```
  * Without calling specific functions, just calling the raster returns class, dimensions, resolution, extent, rs, source, names, and values.

## E6: Check the crs of the raster/nlcd.tif file from the spDataLarge package. What kind of information can you learn from it?
```{r E6, echo=FALSE, warning=FALSE}
crs(nlcd_rast) %>% cat("\n")
```

