---
title: "Chapter 6 Notes"
author: "Darrell A. Carvalho"
output:
  html_document: 
    toc: yes
    toc_depth: 4
    toc_float: yes
    theme: spacelab
    number_sections: yes
---

# Setup

```{r setup, eval = TRUE, include = TRUE}
knitr::opts_chunk$set(echo = TRUE)
# install.packages("remotes")
# remotes::install_github("geocompx/geocompkg")
library(sf)
library(terra)
library(dplyr)
```

# Introduction

-   This chapter covers raster-vector interactions.
    -   The first section covers raster cropping and masking
    -   The second section covers extraction of raster values via vectors
    -   The third and fourth sections cover converting between raster and vector \# Raster Cropping

# Raster Cropping

-   **cropping** reduces the extent of a raster object
-   **masking** sets values outside of a given object/boundary to `NA`

```{r data-import}
srtm <- rast(system.file("raster/srtm.tif", package = "spDataLarge"))
zion <- read_sf(system.file("vector/zion.gpkg", package = "spDataLarge"))
zion <- st_transform(zion, st_crs(srtm))
```

```{r crop}
srtm_cropped <- crop(srtm, zion)
plot(srtm_cropped)
```

```{r mask}
srtm_masked <- mask(srtm, zion)
plot(srtm_masked)
```

-   Often, we wish to both crop and mask, to reduce memory usage and bring a study area into focus

```{r}
srtm_cropped <- crop(srtm, zion)
srtm_final <- mask(srtm_cropped, zion)
plot(srtm_final)
```

-   `mask()` has settings that allow the production of different results
    -   `inverse = TRUE` will invert the masking operation, masking pixels within the mask polygon
    -   `updatevalue = 0` will use 0 as the replacement value, rather than `NA`

```{r inverted-mask}
srtm_inv_masked <- mask(srtm, zion, inverse = TRUE)
plot(srtm_masked)
plot(srtm_inv_masked)
```

# Raster Extraction

## Point Extraction

-   in raster extraction, we identify and return values from a 'target' raster based on 'selector' object(s).
    -   Below, we use `zion_points` to extract elevation from `srtm` raster
    -   we then use `cbind` to bind the elevation values to the `zion_points` geometry, using the matching `ID` column
        -   `cbind` stands for columnar bind

```{r}
data("zion_points", package = "spDataLarge")
elevation <- terra::extract(srtm, zion_points)
zion_points <- cbind(zion_points, elevation) %>% select(-ID)

plot(srtm)
plot(zion_points, pch = 16, add = T)
```

## Linear Extraction (Transect via points)

-   we can also use lines as selectors for extraction
    -   this is not recommended for transect approaches due to measurement challenges
    -   instead, best practice currently is to split the line into points, and extract values along points.

```{r line}
zion_transect <- cbind(c(-113.2, -112.9), c(37.45, 37.2)) %>% 
  st_linestring() %>% 
  st_sfc(crs = crs(srtm)) %>% 
  st_sf(geometry = .)

srtm %>% plot()
zion_transect %>% plot(add = T)
```

-   Above, we have generated a transect line.
-   Next we will give a unique id to the transect
-   Then, we segmentize the line to a set density
-   finally, we cast the segments to point.

```{r}
# give unique ID
zion_transect$id <- 1:nrow(zion_transect)

# segmentize
zion_transect <- st_segmentize(zion_transect, dfMaxLength = 250)

# cast to point
zion_transect <- st_cast(zion_transect, "POINT")

zion_transect
```

-   Now we can determine the distance between points

```{r}
zion_transect <- zion_transect %>% 
  group_by(id) %>% 
  mutate(dist = st_distance(geometry)[, 1])
```

-   Finally, we can extract elevation values on each point

```{r}
library(ggplot2)
zion_elev <- terra::extract(srtm, zion_transect)
zion_transect <- cbind(zion_transect, zion_elev)

plot_transect <- ggplot(zion_transect, aes(as.numeric(dist),
                                           srtm)) +
  geom_line() +
  labs(x = "Distance (m)", y = "Elevation (m a.s.l)") +
  theme_bw()
plot_transect
```

## Polygonal Extraction

-   Returns many values, and can be useful for summary statistics within a polygonally defined zone

```{r}
zion_srtm_values <- terra::extract(x = srtm, y = zion)

# generate summary stats
zion_srtm_values %>%
  group_by(ID) %>% 
  summarize(across(srtm, list(min = min, mean = mean, max = max)))
```

-   similar approaches to the one demonstrated above can be applied to categorical rasters

```{r}
nlcd <- rast(system.file("raster/nlcd.tif", package = "spDataLarge"))
zion2 <- st_transform(zion, st_crs(nlcd))
zion_nlcd <- extract(nlcd, zion2, exact = TRUE)
zion_nlcd %>% 
  group_by(ID, levels) %>% 
  count()
```

# Rasterization

-   Rasterization is the process of generating a raster object from a (set of) vector object(s)
    -   in **terra** we use `rasterize()`
        -   it takes `x` - a vector object to be rasterized
        -   it takes `y` - a template raster object defining extent, resolution, and CRS
            -   resolution is vital - too low, results may miss variability; too high, and compute becomes more burdensome

```{r}
cycle_hire_osm <- spData::cycle_hire_osm
cycle_hire_osm_projected <- st_transform(cycle_hire_osm, "EPSG:27700")
raster_template <- rast(ext(cycle_hire_osm_projected), resolution = 1000,
                        crs = crs(cycle_hire_osm_projected))
```

-   We will try three different rasterization approaches:
    -   First, we will represent presence/absence of cycle hire points
    -   Then, we will use `fun = "length"` to compute a count of points in each cell
    -   Finally, we will compute the `capacity` of each cell using `fun = sum` and `field = "capacity"`

```{r presence-absence}
ch_raster <- rasterize(cycle_hire_osm_projected, raster_template)
ch_raster %>% plot()
```

```{r count}
ch_raster2 <- rasterize(cycle_hire_osm_projected, raster_template, fun = "length")
ch_raster2 %>% plot()
```

```{r capacity}
ch_raster3 <- rasterize(cycle_hire_osm_projected, raster_template, field = "capacity", fun = sum, na.rm = TRUE)
ch_raster3 %>% plot()
```

-   When rasterizing lines, determine whether you want to include pixels touching the lines
    -   the argument `touches` is used for this purpose

```{r california-lines}
california <- dplyr::filter(spData::us_states, NAME == "California")
california_borders <- st_cast(california, "MULTILINESTRING")
raster_template2 <- rast(ext(california), resolution = 0.5, 
                         crs = st_crs(california)$wkt)
```

```{r california-touches}
california_raster1 <- rasterize(california_borders, raster_template2, touches = T)
california_raster1 %>% plot()
```

```{r california-nontouch}
california_raster2 <- rasterize(california_borders, raster_template2, touches = F)
california_raster2 %>% plot()
```

-   For polygons, we also have `touches` argument

```{r california-poly-touch}
calipoly_raster1 <- rasterize(california, raster_template2, touches = T)
calipoly_raster1 %>% plot()
```

```{r california-poly-nontouch}
calipoly_raster1 <- rasterize(california, raster_template2, touches = F)
calipoly_raster1 %>% plot()
```

# Spatial Vectorization

-   Spatial Vectorization is the opposite of rasterization
    -   WARNING! 'vectorization' has a different meaning within R and other programming languages; always use "Spatial Vectorization"
-   Below, we vectorize the centroids of our cells `terra::as.points`

```{r}
elev <- rast(system.file("raster/elev.tif", package = "spData"))
elev_point <- as.points(elev) %>% 
  st_as_sf()
elev %>% plot()
elev_point %>% plot(pch = 19, col = "red", add = T)
```

-   we can create contor lines around phenomena in rasters, as well

```{r}
dem <- rast(system.file("raster/dem.tif", package = "spDataLarge"))

cl <- as.contour(dem) %>% 
  st_as_sf()
plot(dem, axes = FALSE)
plot(cl, add = TRUE)
```

-   vectorizing rasters to polygons can result in blocky polygons; smoothing them (e.g. with **smoothr**) may be more aesthetic, but can fudge numbers in analyses because they alter the spatial coverage from the pixels.

```{r}
grain <- rast(system.file("raster/grain.tif", package = "spData"))
grain_poly <- as.polygons(grain) %>% 
  st_as_sf()
grain_poly %>% plot()
```

# Exercises

```{r}
zion_points_path <- system.file("vector/zion_points.gpkg",
                                package = "spDataLarge")
zion_points <- read_sf(zion_points_path)
srtm <- rast(system.file("raster/srtm.tif", package = "spDataLarge"))
ch <- st_combine(zion_points) %>% 
  st_convex_hull() %>% 
  st_as_sf()
```

## Exercise 1

Crop the `srtm` raster using:

-   `zion_points`

    ```{r}
    crop(srtm, zion_points) %>% plot()
    ```

-   `ch`

    ```{r}
    crop(srtm, ch) %>% plot()

    ```

Are there any differences in the output maps? Next, mask `srtm` using these two datasets. Can you see any difference now? How would you explain it?

```{r}
mask1 <- mask(srtm, zion_points)
mask2 <- mask(srtm, ch)
mask1 %>% plot()
mask2 %>% plot()
```

## Exercise 2

Firstly, extract values from `srtm` at the points represented in `zion_points`. Next, extract average values of `srtm` using a 90 buffer around each point frm `zion_points` and compare these two sets of values. When would extracting values by buffers be more suitable than by points alone?

-   Bonus: Implement extraction using **exactextractr** and compare results.

```{r}
elevation <- terra::extract(srtm, zion_points)
zion_points <- cbind(zion_points, elevation) %>% select(-ID)

# plot(zion_points, pch = 16)

zion_buff <- st_buffer(zion_points, 90)

zion_buff_values <- terra::extract(x = srtm, y = zion_buff, fun = "mean", exact = T)

# generate summary stats
zion_buff_values %>%
  group_by(ID) %>% 
  summarize(across(srtm, mean))
plot(zion_points$srtm, zion_buff_values$srtm)

zion_buff_exact <- exact_extract(srtm, zion_buff, fun = "mean")
plot(zion_buff_values$srtm, zion_buff_exact)
```

## Exercise 3

Subset points higher than 3100 meters in New Zealand (the `nz_height` object) and create a template raster with a resolution of 3 km for the extent of the new point dataset. Using these two new objects:

```{r}
nz_3100 <- spData::nz_height %>% filter(elevation > 3100)
nz_template <- rast(ext(nz_3100), resolution = 3000, 
                         crs = st_crs(nz_3100)$wkt)

```

-   Count numbers of the highest points in each grid cell.

```{r}
nz_raster1 <- rasterize(nz_3100, nz_template, field = "elevation", 
                        fun = "count", na.rm = TRUE)
nz_raster1 %>% plot()
```

-   Find the maximum elevation in each grid cell.

```{r}
nz_raster2 <- rasterize(nz_3100, nz_template, field = "elevation", 
                        fun = "max", na.rm = TRUE)
nz_raster2 %>% plot()
```

## Exercise 4

Aggregate the raster counting high points in New Zealand (created in the previous exercise), reduce its geographic resolution by half (so cells are 6 x 6 km) and plot the result.

```{r}
nz_raster3 <- aggregate(nz_raster1, fact = 2, fun = sum, na.rm = T)
nz_raster3 %>% plot()
```

-   Resample the lower resolution raster back to the original resolution of 3 km. How have the results changed?

```{r}
nz_raster4 <- nz_raster3 %>% 
  resample(nz_template, "near")
nz_raster4 %>% plot()
```

-   Name two advantages and disadvantages of reducing raster resolution.

## Exercise 5

Polygonize the `grain` dataset and filter all squares representing clay.

-   Name two advantages and disadvantages of vector data over raster data.
-   When would it be useful to convert rasters to vectors in your work?

```{r}
grain <- rast(system.file("raster/grain.tif", package = "spData"))
grain_poly <- as.polygons(grain) %>% 
  st_as_sf()

grain_poly %>% filter(grain == "clay") %>% plot()
```
