---
title: "Chapter 6 Notes"
author: "Darrell A. Carvalho"
output:
  html_document: 
    toc: yes
    toc_depth: 4
    toc_float: yes
    theme: spacelab
    number_sections: yes
---

# Setup

```{r setup, eval = TRUE, include = TRUE}
knitr::opts_chunk$set(echo = TRUE)
# install.packages("remotes")
# remotes::install_github("geocompx/geocompkg")
library(sf)
library(terra)
library(dplyr)
```

# Introduction

-   This chapter covers raster-vector interactions.
    -   The first section covers raster cropping and masking
    -   The second section covers extraction of raster values via vectors
    -   The third and fourth sections cover converting between raster and vector \# Raster Cropping

# Raster Cropping

-   **cropping** reduces the extent of a raster object
-   **masking** sets values outside of a given object/boundary to `NA`

```{r data-import}
srtm <- rast(system.file("raster/srtm.tif", package = "spDataLarge"))
zion <- read_sf(system.file("vector/zion.gpkg", package = "spDataLarge"))
zion <- st_transform(zion, st_crs(srtm))
```

```{r crop}
srtm_cropped <- crop(srtm, zion)
plot(srtm_cropped)
```

```{r mask}
srtm_masked <- mask(srtm, zion)
plot(srtm_masked)
```

-   Often, we wish to both crop and mask, to reduce memory usage and bring a study area into focus

```{r}
srtm_cropped <- crop(srtm, zion)
srtm_final <- mask(srtm_cropped, zion)
plot(srtm_final)
```

-   `mask()` has settings that allow the production of different results
    -   `inverse = TRUE` will invert the masking operation, masking pixels within the mask polygon
    -   `updatevalue = 0` will use 0 as the replacement value, rather than `NA`

```{r inverted-mask}
srtm_inv_masked <- mask(srtm, zion, inverse = TRUE)
plot(srtm_masked)
plot(srtm_inv_masked)
```

# Raster Extraction

## Point Extraction

-   in raster extraction, we identify and return values from a 'target' raster based on 'selector' object(s).
    -   Below, we use `zion_points` to extract elevation from `srtm` raster
    -   we then use `cbind` to bind the elevation values to the `zion_points` geometry, using the matching `ID` column
        -   `cbind` stands for columnar bind

```{r}
data("zion_points", package = "spDataLarge")
elevation <- terra::extract(srtm, zion_points)
zion_points <- cbind(zion_points, elevation) %>% select(-ID)

plot(srtm)
plot(zion_points, pch = 16, add = T)
```

## Linear Extraction (Transect via points)

-   we can also use lines as selectors for extraction
    -   this is not recommended for transect approaches due to measurement challenges
    -   instead, best practice currently is to split the line into points, and extract values along points.

```{r line}
zion_transect <- cbind(c(-113.2, -112.9), c(37.45, 37.2)) %>% 
  st_linestring() %>% 
  st_sfc(crs = crs(srtm)) %>% 
  st_sf(geometry = .)

srtm %>% plot()
zion_transect %>% plot(add = T)
```

-   Above, we have generated a transect line.
-   Next we will give a unique id to the transect
-   Then, we segmentize the line to a set density
-   finally, we cast the segments to point.

```{r}
# give unique ID
zion_transect$id <- 1:nrow(zion_transect)

# segmentize
zion_transect <- st_segmentize(zion_transect, dfMaxLength = 250)

# cast to point
zion_transect <- st_cast(zion_transect, "POINT")

zion_transect
```

-   Now we can determine the distance between points

```{r}
zion_transect <- zion_transect %>% 
  group_by(id) %>% 
  mutate(dist = st_distance(geometry)[, 1])
```

-   Finally, we can extract elevation values on each point

```{r}
library(ggplot2)
zion_elev <- terra::extract(srtm, zion_transect)
zion_transect <- cbind(zion_transect, zion_elev)

plot_transect <- ggplot(zion_transect, aes(as.numeric(dist),
                                           srtm)) +
  geom_line() +
  labs(x = "Distance (m)", y = "Elevation (m a.s.l)") +
  theme_bw()
plot_transect
```

## Polygonal Extraction

-   Returns many values, and can be useful for summary statistics within a polygonally defined zone

```{r}
zion_srtm_values <- terra::extract(x = srtm, y = zion)

# generate summary stats
zion_srtm_values %>%
  group_by(ID) %>% 
  summarize(across(srtm, list(min = min, mean = mean, max = max)))
```

-   similar approaches to the one demonstrated above can be applied to categorical rasters

```{r}
nlcd <- rast(system.file("raster/nlcd.tif", package = "spDataLarge"))
zion2 <- st_transform(zion, st_crs(nlcd))
zion_nlcd <- extract(nlcd, zion2, exact = TRUE)
zion_nlcd %>% 
  group_by(ID, levels) %>% 
  count()
```

# Rasterization

-   Rasterization is the process of generating a raster object from a (set of) vector object(s)
    -   in **terra** we use `rasterize()`
        -   it takes `x` - a vector object to be rasterized
        -   it takes `y` - a template raster object defining extent, resolution, and CRS
            -   resolution is vital - too low, results may miss variability; too high, and compute becomes more burdensome

```{r}
cycle_hire_osm <- spData::cycle_hire_osm
cycle_hire_osm_projected <- st_transform(cycle_hire_osm, "EPSG:27700")
raster_template <- rast(ext(cycle_hire_osm_projected), resolution = 1000,
                        crs = crs(cycle_hire_osm_projected))
```

-   We will try three different rasterization approaches:
    -   First, we will represent presence/absence of cycle hire points
    -   Then, we will use `fun = "length"` to compute a count of points in each cell
    -   Finally, we will compute the `capacity` of each cell using `fun = sum` and `field = "capacity"`

```{r presence-absence}
ch_raster <- rasterize(cycle_hire_osm_projected, raster_template)
ch_raster %>% plot()
```

```{r count}
ch_raster2 <- rasterize(cycle_hire_osm_projected, raster_template, fun = "length")
ch_raster2 %>% plot()
```

```{r capacity}
ch_raster3 <- rasterize(cycle_hire_osm_projected, raster_template, field = "capacity", fun = sum, na.rm = TRUE)
ch_raster3 %>% plot()
```

-   When rasterizing lines, determine whether you want to include pixels touching the lines
    -   the argument `touches` is used for this purpose

```{r california-lines}
california <- dplyr::filter(spData::us_states, NAME == "California")
california_borders <- st_cast(california, "MULTILINESTRING")
raster_template2 <- rast(ext(california), resolution = 0.5, 
                         crs = st_crs(california)$wkt)
```

```{r california-touches}
california_raster1 <- rasterize(california_borders, raster_template2, touches = T)
california_raster1 %>% plot()
```

```{r california-nontouch}
california_raster2 <- rasterize(california_borders, raster_template2, touches = F)
california_raster2 %>% plot()
```

-   For polygons, we also have `touches` argument

```{r california-poly-touch}
calipoly_raster1 <- rasterize(california, raster_template2, touches = T)
calipoly_raster1 %>% plot()
```

```{r california-poly-nontouch}
calipoly_raster1 <- rasterize(california, raster_template2, touches = F)
calipoly_raster1 %>% plot()
```

# Spatial Vectorization
