---
title: "Chapter 2 Notes"
author: "Darrell A. Carvalho"
output: 
  html_document: 
    toc: yes
    toc_depth: 4
    toc_float: yes
    theme: spacelab
    number_sections: yes
---

# Setup

The code in the following chunk installs all the packages necessary for reproducing the book.

The authors demontrate loading libraries using the `library()` function, but I prefer using pacman's `p_load()`

```{r setup, eval = TRUE, include = TRUE}
knitr::opts_chunk$set(message = TRUE, warning = TRUE)
# install.packages("remotes")
# remotes::install_github("geocompr/geocompkg")
pacman::p_load(sf, terra, spData, spDataLarge)
```

------------------------------------------------------------------------

# Introduction

## The fundamental geographic data models

-   Vector

    -   Represents the world as points, lines, and polygons

    -   heavily used in social sciences

    -   discrete placement of vector entries leads to high precision

        -   accuracy not guaranteed

-   Raster

    -   aggregates features into equally-sized cells

        -   consistent over space
        -   scalable

-   the book uses sf for vector and terra for raster

-   raster and vector may, if needed, be converted

------------------------------------------------------------------------

# Vector Data

-   based on points defined by coordinates in a Coordinate Reference System

-   Most often 2D (long, lat), possible 3D (long, lat, height; long, lat, time) corresponding to x,y,z coordinate definitions

    -   London may be plotted at `c(-0.1, 51.5)` in a geographic (lon/lat) CRS
    -   It may also be plotted at `c(530000, 180000)` in British National Grid - a projected coordinate system where each unit is 1 meter

-   **sf** gives both classes for vector data, and a command-line for various libraries:

    -   GDAL (geographic I/O)

    -   PROJ (coordinate system manipulation)

    -   GEOS (geometry on projected data)

    -   S2 (spherical geometry)

        -   switching off S2 can be done using `sf::sf_use_s2("FALSE")`

## An introduction to simple features

simple features is open standard developed by Open Geospatial Consortium

-   hierarchical data model

-   supports 18 geometry types

-   7 core types, supported by **sf** (see **Geometry types**)

**sf** supports *s2* which allows for computation on "non-planar" (that is, spherical) data

This chunk demonstrates a class check on the `world` data loaded from **spData**:

```{r class}
class(world)
```

This chunk demonstrates a names check on the columns of the `world` data:

```{r names}
names(world)
```

This chunk demonstrates simple plotting functionality:

```{r plot-vector, warning=FALSE}
plot(world)
```

The base plot plots each column/variable as its own facet. This all operates under the treatment of `sf` objects as dataframes with geographic data stored in spatial columns (in this case, world\$geom)

This chunk shows that `summary()` and similar function calls retain geometry columns - what the authors refer to as 'sticky behavior' - geometry carries along with the data unless deliberately removed.

```{r summary-world}
summary(world["lifeExp"])
```

This chunk demonstrates the ease of subsetting sf data:

```{r subsetting-world}
world_mini <- world[1:2, 1:3]
print(world_mini)
```

Note the inclusion of spatial metadata and the retention of the `geom` column

-   `sf` objects are data frames with of `geom` or `geometry` list columns (class sfc)

-   `sfc` objects are composed of one or more `sfg` objects - `sfg` objects are composed of coordinates

## Why simple features?

Author-listed reasons for sf:

-   Fast read/write

-   Enhanced plotting

-   treated as dataframes in most cases

-   consistent naming of sf functions using `st_` prefix

-   compatibility with %>% pipe

-   tidyverse compatibility

-   ease to convert between sf and sp if needed:

```{r sf-sp}
library(sp)
world_sp <- as(world, "Spatial") # sf to sp

world_sf <- st_as_sf(world_sp) # sp to sf
```

## Basic map making

basic maps are created with `plot()`

-   this defaults to multi-facet plots with one facet for each variable

```{r plot-world-subset}
plot(world[3:6])
```

-   single-variable plotting is easy, however

```{r plot-world-single-var, warning=FALSE}
plot(world["pop"])
```

-   colors set with `col =` argument
-   `add = true` argument layers plot over existing plot(s)

To demonstrate the above, we plot the Asian continent over the world map

```{r plot-asia, warning=FALSE}
world_asia <- world[world$continent == "Asia", ] #subsets asia continent
asia <- st_union(world_asia) # union of all asian polygons into one multipolygon

plot(world["pop"], reset = FALSE) # plot world population, prevent reset of plot key
plot(asia, add = TRUE, col = "red") # plot asian continent over world in red color
```

The authors demonstrate code on modifying aesthetics using `cex` argument (standing for "c"haracter "ex"pansion, I believe)

**Note:** this code does not center the population circles - this is saved for chapter end exercises.

```{r plot-population, warning=FALSE}
plot(world["continent"], reset = FALSE)
cex = sqrt(world$pop) / 10000
world_cents = st_centroid(world, of_largest = TRUE)
plot(st_geometry(world_cents), add = TRUE, cex = cex)
```

## Geometry types

-   POINT

    -   dimensionless geometry, representing coordinates on a surface

-   LINESTRING

    -   geometry with one dimension - length
    -   composed of segments connecting two or more POINTs

-   POLYGON

    -   geometry with two dimensions - length and width

    -   composed of LINESTRING segments that close

        -   that is, the end POINT is coincident with the start POINT

-   MULTIPOINT

    -   a simple feature that carries multiple non-connected POINTs

-   MULTILINESTRING

    -   a simple feature that carries multiple non-connected LINESTRINGs

-   MULTIPOLYGON

    -   a simple feature that carries multiple non-connected POLYGONs

-   GEOMETRY COLLECTION

    -   a simple feature that carries multiple geometries that are non-connected with each other

## The sf class

This section describes the structure of sf objects, but the long and short of it is that sf objects are data frames with spatial information stored in a geometry column.

## Simple feature geometries (sfg)

-   This class represents the geometry types listed above.

-   functions exist to create sfg objects: they follow the format of `st_geometrytype(c(coordinate, coordinate))`

    -   `st_point()`
    -   `st_linestring()`
    -   `st_polygon()`
    -   `st_multipoint()`
    -   `st_multilinestring()`
    -   `st_multipolygon()`
    -   `st_geometrycollection()`

-   the authors demonstrate the creation of different `sfg` objects using vectors, lists, and matrices

## Simple feature columns (sfc)

-   `sfc` objects are a list of `sfg` objects, + CRS info
-   combine geometries with `st_sfc()`
-   set crs with the `crs` argument for `st_sfc()` or `st_sf()`
-   `st_crs()` reports crs of sf object

## The sfheaders package

This section addresses **sfheaders**, which is a faster library for creating and manipulating sf objects. I will reread this section as needed.

## Spherical geometry operations with S2

-   `sf_use_s2()` returns the s2 engine state, and when passed `TRUE` or `FALSE` can change whether the engine is used or not.

```{r plot-s2-india, echo=TRUE}
india <- world[world$name_long == "India", ]
sf_use_s2() # is SF using s2?

india_buffer_with_s2 <-  st_buffer(india, 1)
plot(india_buffer_with_s2[1])
sf_use_s2(FALSE) # turn off s2
india_buffer_without_s2 <- st_buffer(india, 1)
plot(india_buffer_without_s2[1])
sf_use_s2(TRUE)
```

-   There are conditions where switching s2 off is optimal; to do this project-wide, the authors recommend adding `sf::sf_use_s2(FALSE)` to the .RProfile for the project.

# Raster data

-   spatial raster - continuous grid of cells(pixels)

-   most grids are "regular" - each cell has same dimensions

-   data model consists of header and matrix

-   Header defines CRS, extent, origin

    -   Extent is dimensionality - number of rows and columns
    -   origin is "starting point"/ first cell - upper left by default in **terra**

-   raster data can be processed more efficiently and faster than vector data

-   raster cells hold a single value.

-   This value may be numeric or categorical

-   raster useful for continuous phenomena

-   Discrete phenomena can be represented in raster models, but may be better represented in vector.

## R packages for working with raster

three key raster packages

-   **raster** (undergoing mothballing)

-   **terra** - focuses on regular grid

    -    traditional single or multi-band imagery

    -   built on C++ code, uses C++ pointers

    -   has own `SpatVector`object for vectors

    -   relies on built-in functions

-   **stars**

    -   focuses on more exotic grids

    -   Useful for raster data cubes

    -   stores in arrays or just a filepath to raster data

    -   more integrated with sf functions and objects

    -   has own methods built on existing functions

-   converting between `stars` and `terra` objects is easy

    -   **terra** -> **stars**, use `st_as_stars()`

    -   **stars** -> **terra**, use `rast()`

## An introduction to terra

Beginning work with the Zion National Park Data:

```{r terra-intro, warning=FALSE}
raster_filepath <- system.file("raster/srtm.tif", package = "spDataLarge") # store path to data
my_rast <- rast(raster_filepath) # bring data at filepath into r environment
class(my_rast) # ensure data is SpatRaster from terra
print(my_rast) # return raster header
dim(my_rast) # return rows, columns, and layers
ncell(my_rast) # return cell count
res(my_rast) # return spatial resolution
ext(my_rast) # return spatial extent
crs(my_rast) # return coordinate reference system
inMemory(my_rast) # boolean return of whether data is in memory or on disk
```

-   `help("terra-package")` for **terra** functions

## Basic map making

-   **terra** includes `plot()` methods for SpatRaster objects

```{r plot-terra}
plot(my_rast)
```

-   `terra::plotRGB()` creates Red-Green-Blue plot from raster layers
-   **tmap** allows raster/vector static/interactive maps
-   **rasterVis** extends options

## Raster classes

-   `SpatRaster` - **terra** raster objects

```{r terra-objects}
# Step 1: point to location
single_raster_file <- system.file("raster/srtm.tif", package = "spDataLarge")
# step 2: create object with rast()
single_rast <- rast(raster_filepath)
```

-   **terra** uses GDAL drivers.
-   raster header and pointers read into RAM; not usually whole raster
-   `rast()` allows for creating rasters from scratch.

```{r raster-creation}
new_raster <- rast(nrows = 6, ncols = 6, resolution = 0.5,
                    xmin = -1.5, xmax = 1.5, ymin = -1.5, 
                    ymax = 1.5, vals = 1:36)
```

-   `SpatRaster` class allows multiple layers
-   multispectral/multiband imagery
-   raster time-series

```{r multiband-raster}
multi_raster_file <- system.file("raster/landsat.tif", package = "spDataLarge")
multi_rast <- rast(multi_raster_file)
print(multi_rast)
nlyr(multi_rast) # returns number of layers 
```

-   Multi-layered objects can be subset using `terra::subset()`

```{r subset-bands}
multi_rast3 <- subset(multi_rast, 3)
multi_rast4 <- subset(multi_rast, "landsat_4")
```

-   `c()` combines single layers into a multilayer object

```{r combine-bands}
multi_rast34 <- c(multi_rast3, multi_rast4)
```

# Geographic and projected Coordinate Reference Systems

-   Know your data's CRS

    -   is it Geographic or Projected?

    -   **sf** objects are checked with `st_crs()`

    -   **terra** objects are checked with `crs()`

## Geographic coordinate systems

-   Geographic coordinates identify location using lat/long (angular distances)

-   The surface is modeled either as a sphere or an ellipsoid ("oblate spheroid", in my geography courses)

-   Trade-offs:

    -   sphere is simpler

    -   ellipsoid is more accurate; requires two key parameters - equatorial radius - polar radius

-   Two parts of datums

    -   determine which ellipsoid to use
    -   how to describe locations on this ellipsoid using Cartesian Coordinates.

-   datums are geocentric (`WGS84`) or local (`NAD83`)

    -   geocentric are pegged to Earth's center
    -   local are pegged to a surface at a particular location, but may result in inaccuracies at locations further away from the chosen locale of focus.

## Projected coordinate reference systems

-   All projected CRS have a geographic CRS

    -   a projection converts 3D coordinates (long lat) to Easting and Northing (x and y)
    -   linear unit of measurement (meters, feet, miles, kilometers)

-   deformations occur; different projection families preserve different properties:

    -   equal-area projections preserve area
    -   azimuthal projections preserve direction
    -   equidistant preserve distant
    -   conformal preserve local shapes

-   Three main groups of projections:

    -   conic

        -   one or two tangent lines define the cone surface
        -   mid-latitude mapping

    -   cylindrical

        -   one or two tangent lines define the cylindrical surface
        -   global maps

    -   planar/azimuthal

        -   a point or line of tangency defines the planar projection surface
        -   polar maps \# Units

-   CRSs use spatial units for feature relationships and measurements

-   **sf** has native support for units

    -   useful as some CRS use meters, some use feet Units and **units** demonstration:

    ```{r units-demo}
    luxembourg <- world[world$name_long == "Luxembourg",]
    st_area(luxembourg)
    attributes(st_area(luxembourg))

    # use set_units to change units
    units::set_units(st_area(luxembourg), km^2)
    ```

-   be careful to know the units of rasters, as raster packages as yet do not inform you of the units of analysis

    -   remember WGS84 uses Decimal Degrees
    -   remember UTM uses Meters
