---
title: "Chapter 5 Notes"
author: "Darrell A. Carvalho"
output:
  html_document: 
    toc: yes
    toc_depth: 4
    toc_float: yes
    theme: spacelab
    number_sections: yes
---

# Setup

```{r setup, eval = TRUE, include = TRUE}
knitr::opts_chunk$set(echo = TRUE)
# install.packages("remotes")
# remotes::install_github("geocompx/geocompkg")
library(sf)
library(terra)
library(spData)
library(dplyr)
library(spDataLarge)
library(tmap)
```

# Introduction

-   Chapter 5 focuses on geographic/geometric manipulation
    -   5.2 covers 'unary' and 'binary' operations
        -   'unary' - a single geometry in isolation
        -   'binary' - modify one geometry based on another's shape
    -   5.3 covers geometric operations on rasters
        -   size change
        -   resolution change

# Geometric operations on vector data

## Simplification

-   the process of generalization of vector objects
    -   for smaller scale maps
    -   reduce memory, disk space, and/or bandwith consumed
-   **sf** provides `st_simplify()` for vertex count reduction
    -   `dTolerance` controls generalization level in map units

### Simplification of a `linestring`

```{r}
seine_simp <- st_simplify(seine, 
                          dTolerance = 2000, 
                          preserveTopology = T) # 2000 m
plot(seine)
plot(seine_simp)
```

## Simplification of polygons

```{r}
us_states_simp1 <- st_simplify(us_states, 
                               dTolerance = 100000, preserveTopology = T) # 100 km
plot(us_states$geometry)
plot(us_states_simp1$geometry)
```

```{r}
us_states_simp2 <-  rmapshaper::ms_simplify(us_states, keep = 0.01,
                                          keep_shapes = TRUE)
plot(us_states$geometry)
plot(us_states_simp2$geometry)
```

```{r}
us_states_simp3 <- smoothr::smooth(us_states, 
                                   method = "ksmooth", 
                                   smoothness = 6)

plot(us_states_simp3$geometry)
```

## Centroids

-   centroid operations identify a geographic object's central point
    -   this center can be defined in many ways
    -   in all cases, this reduces a geographic object to a single point feature
-   The most common centroid operation is the *geographic centroid*
    -   represents center of mass in an object
    -   this is computed using `sf::st_centroid()`

```{r warning=FALSE}
nz_centroid <- st_centroid(nz)
seine_centroid <- st_centroid(seine)
plot(nz$geom, border = "green")
plot(nz_centroid, add = T, col = "black")

plot(seine$geometry, col = "blue")
plot(seine_centroid, add = T, col = "black")
```

-   In objects with holes, the centroid can fall external to the object
    -   in these instances, we can use *point on surface* operations to ensure we are getting a centroid on the object's surface

```{r warning=FALSE}
nz_pos <- st_point_on_surface(nz)
seine_pos <- st_point_on_surface(seine)

plot(seine$geometry, col = "grey")
plot(seine_pos, add = T, col = "red")

plot(nz$geom, border = "grey")
plot(nz_pos, add = T, col = "red")
```

## Buffers

-   buffers are polygons showing area-within-distance of a given geometry
    -   buffers are useful for geographic data analysis
        -   e.g. "how many points are within 50 km of line?"
-   `st_buffer()` requires (at least) two arguments
    -   input geometry
    -   distance to compute (in units of CRS)

```{r}
seine_buff_5km <- st_buffer(seine, dist = 5000)
seine_buff_50km <- st_buffer(seine, dist = 50000)


plot(seine_buff_50km$geometry, col = sf.colors(3))
plot(seine_buff_5km, add = T)
plot(seine$geometry, add = T)
```

### Important arguments for `st_buffer()`

-   `x` = the object being buffered
-   `dist` = the distance, in CRS units, to be buffered around the object
-   `nQuadSegs` (when GEOS in use) = number of segments per quadrant
    -   defaults to 30; each circle is composed of 4 quadrants - 30 segments per quadrants means 120 segments per computed circle
    -   set to `FALSE` when using `sf_use_s2()` or when using projected CRS
-   `max_cells` (when S2 engine in use) = maximum number of cells to use to approximate a buffer
    -   As `max_cells` increases, smoothness of buffer increases, but so does computation time
        -   see `?s2::s2_buffer_cells` for details
-   `endCapStyle`/`joinStyle` (when GEOS in use) = controls the appearance of the buffer's edges
-   `singleSide` (when GEOS engine is used) = controls whether buffer is on one or both sides of input

## Affine transformations

-   An *affine transformation* is any transformation that preserves lines and parallelism
    -   angles and lengths may be distorted
    -   common affine transformations include:
        -   shift/translation - the movement of every point by the same distance in map units
        -   scale - enlargement or shrinkage of an object by a stable factor, either globally or locally
            -   global scaling increases or decreases all coordinate values in relation to origin coordinates
                -   maintains topological relations
                -   performed by multiplication or subtraction of an `sfg` or `sfc` object
            -   local scaling operates on geometries independently
                -   requires points to scale geometries around, usually centroids
    -   rotation in two-dimensional space requires a rotation matrix:
        -   The below matrix rotates points in a clockwise direction

$$R = \begin{bmatrix}
\cos \theta & -\sin \theta \\ 
\sin \theta & \cos \theta \\ 
\end{bmatrix}
$$

-   **sf** performs affine transformations on objects of the `sfg` and `sfc` classes

```{r}
nz_sfc <- st_geometry(nz)

# Shifting
nz_shift = nz_sfc + c(0, 100000)
plot(nz_sfc, main = "Shift")
plot(nz_shift, add = T, border = "red")

# Local Scaling
nz_centroid_sfc <- st_centroid(nz_sfc)
nz_scale <- (nz_sfc - nz_centroid_sfc) * 0.5 + nz_centroid_sfc
plot(nz_sfc, main = "Local Scaling")
plot(nz_scale, add = T, border = "red")

# Implementing rotation matrix in r
rotation <- function(a){
  r = a * pi / 180 # degrees to radians
  matrix(c(cos(r), sin(r), -sin(r), cos(r)), nrow = 2, ncol = 2)
}
# our rotation function accepts one argument - rotation angle in degrees

# Rotation
nz_rotate <- (nz_sfc - nz_centroid_sfc) * rotation(30) + nz_centroid_sfc

plot(nz_sfc, border = "black", main = "Rotation")
plot(nz_rotate, border = "red", add = T)
```

-   these newly generated geometries can replace the previous ones via `st_set_geometry`

```{r}
nz_scale_sf <- st_set_geometry(nz, nz_scale)
plot(nz_scale_sf)
```

## Clipping

-   Spatial clipping is spatial subsetting involving changes to `geometry` columns
    -   only geometries more complex than points can be clipped

```{r}
b <- st_sfc(st_point(c(0,1)), st_point(c(1,1))) # create two points
b = st_buffer(b, dist = 1) # create circles from the points
plot(b, border = "gray")
text(x = c(-0.5, 1.5), y = 1, labels = c("x", "y"), cex = 3) # add text
```

```{r}
# clipping to intersect
x <- b[1]
y <- b[2]
x_and_y <- st_intersection(x, y)
plot(b, border = "gray")
plot(x_and_y, col = "lightgray", border = "gray", add = TRUE) # intersecting area

# clipping to only area in X but not Y
x_not_y <- st_difference(x, y)
plot(b, border = "gray")
plot(x_not_y, col = "lightgray", border = "gray", add = TRUE)

# clipping to only area in Y but not X
y_not_x <- st_difference(y, x)
plot(b, border = "gray")
plot(y_not_x, col = "lightgray", border = "gray", add = TRUE)

# merging x and y
x_union_y <- st_union(x, y)
plot(b, border = "gray")
plot(x_union_y, col = "lightgray", border = "gray", add = TRUE)

# clipping to include x and y but not the intersection of x and y
x_y <- st_sym_difference(x, y)
plot(b, border = "gray")
plot(x_y, col = "lightgray", border = "gray", add = TRUE)
```

## Subsetting and clipping

-   Clipping may change geometries but may also subset objects returning only features that intersect with a subsetting object

```{r}
# create a bounding box around the union of x and y
bb <- st_bbox(st_union(x, y))
box <- st_as_sfc(bb)

# set seed
set.seed(2024)

# select 10 random points from within the bounding box
p <- st_sample(x = box, size = 10)
# use bracket subsetting to select only the point(s) within the intersection of x and y
p_xy1 <- p[x_and_y]

# plot it all out
plot(box, border = "gray", lty = 2)
plot(x, add = T, border = "gray")
plot(y, add = TRUE, border = "gray")
plot(p, add = TRUE, cex = 3.5)
plot(p_xy1, cex = 5, col = "red", add = T)
text(x = c(-0.5, 1.5), y = 1, labels = c("x", "y"), cex = 3)
```

-   We can perform similar results using the code chunks below
    -   as above, `p_xy1` is using an intermediate option and brackets to find a subset
    -   `p_xy2` uses `st_intersection` to find the intersection between the points and our `x_and_y` object
        -   This will return features that partially intersect, but modifies their geometries where they cross the borders
    -   `p_xy3` uses the binary predicate `st_intersects()` to create a subsetting object which is then passed to bracketing

```{r}
# create a bounding box around the union of x and y
bb <- st_bbox(st_union(x, y))
box <- st_as_sfc(bb)

# set seed
set.seed(2024)

# select 10 random points from within the bounding box
p <- st_sample(x = box, size = 10)

# way 1
p_xy1 <- p[x_and_y]

# way 2
p_xy2 <- st_intersection(p, x_and_y)

# way 3
sel_p_xy <- st_intersects(p, x, sparse = FALSE)[, 1] &
  st_intersects(p, y, sparse = FALSE) [, 1]
p_xy3 <- p[sel_p_xy]
```

-   all of the above approaches create effectively identical results
    -   unless other factors intervene, prefer the simplest/most concise implementation

## Geometry Unions

-   spatial aggregation can silently dissolve geometries of touching polygons with same grouping

```{r}
regions <- aggregate(x = us_states[, "total_pop_15"], 
                     by = list(us_states$REGION),
                     FUN = sum, na.rm = TRUE)
regions <- us_states %>% 
  group_by(REGION) %>% 
  summarize(pop = sum(total_pop_15, na.rm = TRUE))
palette = c("lightblue", "blue4")
qtm(regions, fill = "pop", 
    fill.scale = tm_scale_intervals(n = 4,
                                    style = "equal",
                                    values = rev(palette)))
```

-   Under the hood, R uses `aggregate()` and `summarize()` to combine the geometries
-   then, `st_union()` is used to dissolve boundaries and create unified geometries.
    -   in terms of geometries, it is effectively running code akin to below for each region

```{r}
us_west <- us_states[us_states$REGION == "West", ]
us_west_union <- st_union(us_west)
plot(us_west_union)
```

```{r}
texas <- us_states[us_states$NAME == "Texas", ]
texas_union <- st_union(us_west_union, texas)
plot(texas_union)
```

## Type transformations

-   transformation of geometry type in R is referred to as *geometry casting*
    -   **sf**'s `st_cast()` is used for geometry casting
        -   it presents different behaviors when operating on a single `sfg`, on a simple feature column (`sfc`), and on **sf** objects

```{r}
multipoint <- st_multipoint(matrix(c(1, 3, 5, 1, 3, 1), ncol = 2))
```

-   given a simple feature geometry object (`sfg`), `st_cast()` can readily transform the object between types
    -   in our case, from **multipoint** to **linestring** or **polygon**

```{r}
linestring <- st_cast(multipoint, "LINESTRING")
polyg <- st_cast(multipoint, "POLYGON")
multipoint %>% plot()
linestring %>% plot()
polyg %>% plot()
```

-   `st_cast()` operations can also be performed in the opposing direction, 'reversing' the operations

```{r}
multipoint_2 <- st_cast(linestring, "MULTIPOINT")
multipoint_3 <- st_cast(polyg, "MULTIPOINT")
all.equal(multipoint, multipoint_2)
all.equal(multipoint, multipoint_3)
```

-   When casting on `sfg` objects, one can cast from non-multi-types to multi-types, and the inverse
    -   when casting from multi-types to non-multi-types, only the first element remains
-   casting on `sfc` objects and `sf` objects work similarly, with some caveats
    -   conversion between multi-types and non-multi-types result into many non-multi-objects

```{r}
multilinestring_list <- list(matrix(c(1, 4, 5, 3), ncol = 2),
                       matrix(c(4, 4, 4, 1), ncol = 2),
                       matrix(c(2, 4, 2, 2), ncol = 2))

multilinestring <- st_multilinestring(multilinestring_list)
multilinestring_sf <- st_sf(geom = st_sfc(multilinestring))
multilinestring_sf
multilinestring_sf %>% plot()
```

-   imagine the above is a road network
    -   it is a single row that defines all lines
        -   you are unable to name each linestring by street name
        -   you are unable to calculate the length of a given segment
    -   we can use `st_cast()` to separate the multilinestring into three linestrings

```{r}
linestring_sf2 <- st_cast(multilinestring_sf, "LINESTRING")
linestring_sf2
```

-   With this newly created object, we can now create attributes and measurements for the segments

```{r}
linestring_sf2$name = c("Riddle Rd", "Marshall Ave", "Foulke St")
linestring_sf2$length = st_length(linestring_sf2)
linestring_sf2
```

# Geometric operations on raster data

-   The following are all geometric operations which are useful with raster data:
    -   shifting
    -   flipping
    -   mirroring
    -   scaling
    -   rotation
    -   warping
-   all of these operations contribute to vital tasks like georeferencing - the overlaying of rasters on maps with a known, often different, CRS
-   Georeferencing techniques include
    -   georectification using ground control points
    -   orthorectification, including local topography
    -   image registration (combining images of the same object from different sensors via image alignment)
-   R is not conducive to geo- or orthorectification
-   R is excellent for image registration

## Geometric intersections

-   subsetting a raster and retrieving its spatial output is similar to value-extraction in R
    -   the key difference is that we must use `drop = FALSE` to get our raster object

```{r}
elev <- rast(system.file("raster/elev.tif", package = "spData"))
clip <- rast(xmin = -0.5, xmax = 0.5, ymin = -0.5, ymax = 0.5,
             resolution = 0.3, vals = rep(1, 9))
elev %>% plot()
clip %>% plot()
elev[clip, drop = FALSE] %>% plot()
```

-   we could get the same result `crop()`

```{r}
elev <- rast(system.file("raster/elev.tif", package = "spData"))
clip <- rast(xmin = -0.5, xmax = 0.5, ymin = -0.5, ymax = 0.5,
             resolution = 0.3, vals = rep(1, 9))
elev %>% plot()
clip %>% plot()
elev[clip, drop = FALSE] %>% plot()
crop(elev, clip) %>% plot()
```

## Extent and Origin

-   map algebra requires that resolution, projection, origin and/or extent match between rasters
    -   We can adjust mismatches through alignment
    -   in the simplest case, images differ by extent.
        -   In such cases, the extent of one image can be made to match the other by adding null pixels around the raster such that their extent maches
        -   in **terra** this can be done using `extend()`

```{r}
elev <- rast(system.file('raster/elev.tif', package = 'spData'))
elev_2 <- extend(elev, c(1, 2))

elev_4 <- extend(elev, elev_2)
```

-   The origin (cell corner closest to coordinates (0, 0)) of a raster can be checked or changed using `origin()`

```{r}
origin(elev_4)
elev_4 %>% plot()
origin(elev_4) <- c(0.25, 0.25)
elev_4 %>% plot(add = T)
```

## Aggregation and Disaggregation

-   decreasing a raster's resolution is done via `aggregate()`
    -   when aggregating, a function is supplied to determine how to compute aggregated cells
-   increasing a raster's resolution is done via `disagg()`
    -   there are two methods included in `disagg()`
        -   `"near"` gives output cells the value of the input cells, duplicating values
        -   `"bilinear"` uses the four nearest pixel centers of input to compute average (weighted by distance) for output
        -   Note that disaggregation does not reverse aggregation; the values are only as good as their lowest resolution source.

```{r}
dem <- rast(system.file("raster/dem.tif", package = "spDataLarge"))
dem_agg <- aggregate(dem, fact = 5, fun = mean)
dem_agg2 <- aggregate(dem, fact = 5, fun = median)
plot(dem)
plot(dem_agg)
plot(dem_agg2)
```

```{r}
dem_disagg <- disagg(dem_agg, fact = 5, method = "bilinear")
identical(dem, dem_disagg) # the dem and the disaggregated dem_agg are not identical
plot(dem_agg)
plot(dem)
plot(dem_disagg)
```

## Resampling

-   when we have two or more rasters with differing resolution ans origins, **resampling** comes into play
    -   **resampling**: the process of computing values for new pixel locations
        -   we take values of an original raster and recalculate new values for a target raster with custom resolution and origin
    -   There are many resampling methods:
        -   Nearest neighbor - assign the value of the nearest cell of the original raster to the new /target cell
        -   Bilinear interpolation - assign weighted average of four nearest cells from original to new/target
        -   Cubic interpolation - uses values of the 16 nearest cells of original to new/target, using third-order polynomials
            -   Cubic spline interpolation - also uses 16 values, but uses cubic splines (piece-wise third-order
        -   Lanczos windowed sinc resampling: uses 36 nearest cells to determine new/target cell value
    -   We will use `terra::resample()` for our purposes
        -   `resample` accepts an input raster (`x`), target spatial properties (`y`), and a `method`

```{r}
# Create a sample raster to demonstrate on
target_rast <- rast(xmin = 794650, xmax = 798250,
                    ymin = 8931750, ymax = 8935350,
                    resolution = 300, crs = "EPSG:32717")

dem_resample  <-  resample(dem, y = target_rast, method = "cubicspline")
plot(dem)
plot(dem_resample)
```

# Exercises

E1. Generate and plot simplified versions of the `nz` dataset. Experiment with different values of `keep` (ranging from 0.5 to 0.00005) for `ms_simplify()` and `dTolerance` (from 100 to 100,000) for `st_simplify()`.

```{r}
nz$geom %>% rmapshaper::ms_simplify(keep = 0.05) %>% plot()

nz$geom %>% st_simplify(dTolerance = 10000) %>% plot()
```

E2. In the first exercise in Chapter Spatial Data Operations, it was established that Canterbury region had 70 of the 101 highest points in New Zealand. Using `st_buffer()`, how many points in `nz_height` are within 100 km of Canterbury?

```{r}
cant_buff <- nz[nz$Name == "Canterbury",] %>% st_buffer(100000)
cant_buff$geom %>% plot(col = "grey")
nz$geom %>% plot(col = "white" ,add = T)
nz[nz$Name == "Canterbury",]$geom %>% plot(col = "green", add = T)
nz_height$geometry[cant_buff] %>% plot(add = T)
nz_height$geometry[cant_buff] %>% mapview::npts()
```

E3. Find the geographic centroid of New Zealand. How far is it from the geographic centroid of Canterbury?

```{r}
nz_cent <- nz$geom %>% st_union() %>% st_centroid()
cant_cent <- nz[nz$Name == "Canterbury",]$geom %>% st_centroid()

dist_line <- st_union(nz_cent, cant_cent, by_feature = TRUE) %>% st_cast("LINESTRING")

nz$geom %>% st_union %>% plot()
dist_line %>% plot(add = T, col = "grey")
cant_cent %>% plot(add = T, col = "blue")
nz_cent %>% plot(add = T, col = "red")
st_distance(nz_cent, cant_cent) %>% units::set_units("km")
st_length(dist_line) %>% units::set_units("km")
```

E4. Most world maps have a north-up orientation. A world map with a south-up orientation could be created by a reflection of the `world` object geometry. Write code to do so. Hint: you can use the `rotation()` function from this chapter for this transformation.

```{r}
rotation <- function(a){
  r = a * pi / 180 # degrees to radians
  matrix(c(cos(r), sin(r), -sin(r), cos(r)), nrow = 2, ncol = 2)
}
world_sfc <- st_geometry(world)
world_rotate <- world_sfc * rotation(180)
world_rotate <- st_set_geometry(world, world_rotate)
world_rotate %>% plot()
```

Bonus: create an upside down map of your country.

```{r}
rotation <- function(a){
  r = a * pi / 180 # degrees to radians
  matrix(c(cos(r), sin(r), -sin(r), cos(r)), nrow = 2, ncol = 2)
}
us <- world[world$name_long == "United States",]
us_sfc <- st_geometry(us)
us_rotate <- us_sfc * rotation(180)
us_rotate <- st_set_geometry(us, us_rotate)
us_rotate %>% plot()
```

E5. Run the code in Section 5.2.6. With reference to the objects created in that section, subset the point in `p` that is contained within `x` AND `y`.

```{r}
b <- st_sfc(st_point(c(0,1)), st_point(c(1,1))) # create two points
b = st_buffer(b, dist = 1) # create circles from the points
plot(b, border = "gray")
text(x = c(-0.5, 1.5), y = 1, labels = c("x", "y"), cex = 3) # add text

# clipping to intersect
x <- b[1]
y <- b[2]

# set seed
set.seed(2024)

# select 10 random points from within the bounding box
p <- st_sample(x = box, size = 10)

# using bracket/base subsetting
p_in_y <- p[y]
p_in_xy <- p_in_y[x]

# using st_intersection() to create an intermediate object
x_and_y <- st_intersection(x, y)
p_in_x_and_y <- p[x_and_y]

# Plot to view
plot(box, border = "gray", lty = 2)
plot(x, add = T, border = "gray")
plot(y, add = TRUE, border = "gray")
plot(p, add = TRUE, cex = 3.5)
plot(p_in_x_and_y, cex = 5, col = "red", add = T)
text(x = c(-0.5, 1.5), y = 1, labels = c("x", "y"), cex = 3)

identical(p_in_xy, p_in_x_and_y)
```

-   using base subsetting operators

```{r}
# using bracket/base subsetting
p_in_y <- p[y]
p_in_xy <- p_in_y[x]
# Plot to view
plot(box, border = "gray", lty = 2)
plot(x, add = T, border = "gray")
plot(y, add = TRUE, border = "gray")
plot(p, add = TRUE, cex = 3.5)
plot(p_in_xy, cex = 5, col = "red", add = T)
text(x = c(-0.5, 1.5), y = 1, labels = c("x", "y"), cex = 3)

```

-   Using an intermediary object created with the `st_intersection()`

```{r}
x_and_y <- st_intersection(x, y)
p_in_x_and_y <- p[x_and_y]
plot(box, border = "gray", lty = 2)
plot(x, add = T, border = "gray")
plot(y, add = TRUE, border = "gray")
plot(p, add = TRUE, cex = 3.5)
plot(p_in_x_and_y, cex = 5, col = "red", add = T)
text(x = c(-0.5, 1.5), y = 1, labels = c("x", "y"), cex = 3)
```

-   Let's see if they're identical

```{r}
identical(p_in_xy, p_in_x_and_y)
```

E6: Calculate the length of the boundary lines of US states in meters. Which state has the longest border and which has the shortest?

```{r}
us_states9311 <- st_transform(us_states, "EPSG:9311")
us_states_bor <- st_cast(us_states9311, "MULTILINESTRING")
us_states_bor$borders <- st_length(us_states_bor)
us_states_bor %>% select(NAME, borders) %>% arrange(borders) %>% slice(-(2:(n()-1)))
```

E7 Read the srtm.tif file into R. This raster has a resolution of 0.00083 \* 0.00083 degrees. Change its resolution to 0.01 \* 0.01 degrees using all the methods available in terra package. Visualize the results. Can you notice any differences between the results of these resampling methods?

```{r}
srtm <- rast(system.file("raster/srtm.tif", package = "spDataLarge"))
target_rast <- rast(xmin = -113.2396, xmax = -112.8521,
                    ymin = 37.13208, ymax = 37.51292,
   
                 resolution = 0.01, crs = "EPSG:4326")
```

```{r}
# Original
srtm %>% plot(main = "Original")

# Nearest Neighbor
nearneigh <- resample(srtm, target_rast, "near")
nearneigh %>% plot(main = "Nearest Neighbor")

# Bilinear
bilin <- resample(srtm, target_rast, "bilinear")
bilin %>% plot(main = "Bilinear Interpolation")
# Cubic
cubic <- resample(srtm, target_rast, "cubic")
cubic %>% plot(main = "Cubic Interpolation")

# Cubic Spline
cspline <- resample(srtm, target_rast, "cubicspline")
cspline %>% plot(main = "Cubic B-spline Interpolation")

# Lanczos Windowed Sinc Resampling
lanczo <- resample(srtm, target_rast, "lanczos")
lanczo %>% plot(main = "Lanczos Windowed Resampling")
```
