---
title: "Chapter 4 Notes"
author: "Darrell A. Carvalho"
output:
  html_document: 
    toc: yes
    toc_depth: 4
    toc_float: yes
    theme: spacelab
    number_sections: yes
---

# Setup

```{r setup, eval = TRUE, include = TRUE}
knitr::opts_chunk$set(echo = TRUE)
# install.packages("remotes")
# remotes::install_github("geocompx/geocompkg")
pacman::p_load(sf, terra, spData, dplyr)
elev = rast(system.file("raster/elev.tif", package = "spData"))
grain = rast(system.file("raster/grain.tif", package = "spData"))
```

# Introduction

-   Chapter demonstrates **spatial operations** or "how spatial objects can be modified ... based on their location and shape."
    -   Included are **spatial subsetting**, **spatial joining**, and **spatial aggregation**
-   **Spatial joins** include a variety of means for joining - from **intersect** to **within distance** operations, as all spatial objects have a **spatial relationship**.
    -   distance calculations help us determine the strength and/or relevance of this relationship.
    
-   **Spatial operations** on raster objects:
    -   **subsetting**
    -   **merging**
    -   **Map Algebra**
        -   **local**
        -   **focal**
        -   **zonal**
        -   **global**

# Spatial operations on vector data
## Spatial Subsetting
-   **Spatial subsetting**: "process of taking a spatial object and returning ... only features that *relate* in space to another object."
    -   can be created with square brackets `[`
        -   syntax: `x[y, , op = st_intersects]`
            -   **x** is an `sf` object to be subset 
            -   **y** is the 'subsetting object'
            -   `op = st_intersects` is an optional topological argument
    -   Demonstration of spatial subsetting using `nz` and `nz_height`
```{r canterbury-object}
canterbury <- nz |> filter(Name == "Canterbury")
canterbury_height <- nz_height[canterbury, ]
# nz_height[canterbury, , op = st_disjoint] %>% plot()
```
-   **spatial subsetting** can also be performed with **topological operators**
    -   the code chunk below returns a **sparse geometry binary predicate**
        -   `sgbp` are *sparse matrices*, lists of integer vectors with ordered, TRUE indices for each row.
        -   in other words, it does not store the entire matrix, instead storing only non-zero values and their location within a matrix, thus removing zero-values and their storage requirements.
```{r topological-operator}
sel_sgbp <- st_intersects(nz_height, canterbury)
sel_sgbp
```
    - This chunk can then be used to subset
```{r topological-operator-subsetting}
sel_logical <- lengths(sel_sgbp) > 0
canterbury_height2 <- nz_height[sel_logical, ]
```
```{r topological-operator-subetting-with-st_filter}
canterbury_height3 <- nz_height |>
  st_filter(y = canterbury, .predicate = st_intersects)
```

## Topological relations
-   A topological relationship defines spatial relations between objects.
  -    **Binary topological relationships** are logical relations (`T` or `F`)
  -   these logic statements are made about two objects
      -   the objects are defined by ordered sets of points that define the points, lines, and polygons.
- `sf` has 'binary predicates' - functions that test topological relations
```{r topological-relations-demo}
# Create a polygon
polygon_matrix <- cbind(
  x = c(0,0,1,1,0),
  y = c(0,1,1,0.5,0)
)
polygon_sfc <- st_sfc(st_polygon(list(polygon_matrix)))

# create line
line_sfc <- st_sfc(st_linestring(cbind(
  x = c(0.4, 1),
  y = c(0.2, 0.5)
)))

# create points
point_df <- data.frame(
  x = c(0.2, 0.7, 0.4),
  y = c(0.1, 0.2, 0.8)
)

point_sf <- st_as_sf(point_df, coords = c("x", "y"))
```

```{r predicate-checks}
st_intersects(point_sf, polygon_sfc) # do any of the points intersect the poly?
st_within(point_sf, polygon_sfc) # are any of them within the polygon?
st_touches(point_sf, polygon_sfc) # do any points touch the polygon?
st_disjoint(point_sf, polygon_sfc) # are any of the points exist outside poly?
st_is_within_distance(point_sf, polygon_sfc, dist = 0.2) # are points w/i 0.2 unit distance of poly?
```
## DE-9IM Strings
## Spatial Joining
-   Spatial Joins use spatial predicates as keys, unlike non-spatial joins
```{r reprex}
set.seed(2018) # set seed for reproducibility
(bb = st_bbox(world)) # the world's bounds
#>   xmin   ymin   xmax   ymax 
#> -180.0  -89.9  180.0   83.6
random_df = data.frame(
  x = runif(n = 10, min = bb[1], max = bb[3]),
  y = runif(n = 10, min = bb[2], max = bb[4])
)
random_points = random_df |> 
  st_as_sf(coords = c("x", "y"), crs = "EPSG:4326") # set coordinates and CRS

world_random = world[random_points, ]
nrow(world_random)
#> [1] 4
random_joined = st_join(random_points, world["name_long"])
```

```{r}
tmap::qtm(random_points)
tmap::qtm(world_random)
tmap::qtm(random_joined)
```

