---
title: "Chapter 1 Notes"
author: "Darrell A. Carvalho"
date: "1/7/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
* Book focuses on spatial skills
  + geographic IO
  + data manipulation
  + static and interactive mapping
  + spatial modelling

## What is geocomputation?
* **Geocomputation:** "working with geographic data in a computational way, focusing on code, reproducibility and modularity."
  + Authors trace it to 1996 conference
  + Differentiated from 'quantitative geography' by focus on experimentation and innovation
  + early advocates identified as Longley, Openshaw, Abrahart
  + Robin et al. emphasize "reproducibility an collaboration" in their vision of Geocomputation

## Why use R for Geocomputation?
* "advanced analysis, modelling, and visualization capabilities"
* User friendly due to RStudio
* a lot of "bridges" to other tools/languages
* rapidly evolving - authors demonstrate with following code chunk, which wouldn't have been as easy to implement in 2013

```{r}
library(leaflet)
popup = c("Robin", "Jakub", "Jannes")
leaflet() %>%
  addProviderTiles("NASAGIBS.ViirsEarthAtNight2012") %>%
  addMarkers(lng = c(-3, 23, 11),
             lat = c(52, 53, 49), 
             popup = popup)
```

## Software for geocomputation
* Interpreted languages
  + Python
      - QGIS, GRASS, SAGA have Python APIs
      - popular in data science
      - general purpose language
      - large community across domains
  + R
      - better support of vector and raster in language
      - strong visualization tools
      - focused on statistical/data applications

* Compiled languages
  + C++
      - basis of QGIS, GRASS, SAGA
  + Java
      - basis for GeoTools, JTS, Geoserver/Geonode
      - platform independent

## R's spatial ecosystem
* rapidly evolving
  + **sf** builds off of **sp**

* influence of **dplyr** and **tidyverse**
  + **sf** creates rough approach to 'tidy geographic data'

* rspatial and r-spatial are two related but separate communities
  + **rspatial** is responsible for **terra**
  + **r-spatial** is responsible for **sf**

* earlier packages
  + **akima** and **geoR** - smoothing and interpolation
  + **splancs** and **spatstat** - point pattern analysis
  + **spdep** - spatial autocorrelation
  + **maptools** - allowed reading of shapefiles into R; precursor to **sp**
  + **sp** - allowed for distinctly spatial classes and methods instead of treating spatial data the same as non-spatial data
  + **rgdal** - allowed the import of a variety of raster (and through PROJ, vector) spatial data
  + **rgeos** - interfaces with GEOS, allowing geometric operations, mainly on **sp** data
  + **raster** - allowed raster support not available in **sp**, data too large for RAM, and implemented map algebra

* More recent developments
  + Various bridge packages (**GRASS, spgrass6, and rgrass7** to GRASS;
  **RSAGA** to SAGA; **rqgisprocess** to QGIS)
  + expansion of visualization
      - **RGoogleMaps** allowed pulling in online basemaps
      - **ggmap** allowed for inefficient mapping via **ggplot2**
      - **raster** was expanded on by **rasterVis**
  + 2018 - development of **tmap**, **leaflet**, and **mapview** as 
  visualization packages that support **sf** data
  + 2018 - switch from **proj4string** to **WKT** due to changes in PROJ
  + 2018 - **rayshader** package gives raytracing and hill-shading tools
  + 2018 - **ggspatial** adds visualization elements to **ggplot2**
  + 2019 - **gganimate** opens the door for smooth animation within **ggplot2**
  + 2020 - **terra** is developed as a successor to **raster** 
  + 2021 - **sf** incorporates spherical geometry
      - operations on geographic CRS use *s2geometry* from C++
      - operations on projected CRS still use *GEOS*
  + 2021 - **stars** developed to handle data cubes
  + 2021 - **lidR** processes LiDAR data
  + 2021 - the announced retirement of **rgdal**, **rgeos**, and **maptools**
  
