---
title: "Chapter 2 Notes"
author: "Darrell A. Carvalho"
output: html_document
---

## Setup

The code in the following chunk installs all the packages necessary for 
reproducing the book.

```{r setup, eval = FALSE, include = TRUE}
install.packages("remotes")
remotes::install_github("geocompr/geocompkg")
```

***

The authors demontrate loading libraries using the `library()` function, 
but I prefer using pacman's `p_load()`

```{r libraries, eval=FALSE, include=TRUE}
pacman::p_load(sf, terra, spData, spDataLarge)
```

***

## Introduction

#### **The fundamental geographic data models:**

* Vector
  + Represents the world as points, lines, and polygons
  + heavily used in social sciences
  + discrete placement of vector entries leads to high precision
      - accuracy not guaranteed
* Raster
  + aggregates features into equally-sized cells
      - consistent over space
      - scaleable
* the book uses sf for vector and terra for raster
* raster and vector may, if needed, be converted

***

## Vector Data

* based on points defined by coordinates in a Coordinate Reference System
* Most often 2D (long, lat), possible 3D (long, lat, height; long, lat, time)
corresponding to x,y,z coordinate definitions
  + London may be plotted at `c(-0.1, 51.5)` in a geographic (lon/lat) CRS
  + It may also be plotted at `c(530000, 180000)` in British National Grid - a
  projected coordinate system where each unit is 1 meter
  
* **sf** gives both classes for vector data, and a command-line for various
libraries:
  + GDAL (geographic I/O)
  + PROJ (coordinate system manipulation)
  + GEOS (geometry on projected data)
  + S2 (spherical geometry)
      - switching off S2 can be done using `sf::sf_use_s2("FALSE")`

#### **An introduction to simple features**
* simple features is open standard developed by Open Geospatial Consortium
  + hierarchical data model
  + supports 18 geometry types
  + 7 core types, supported by **sf**:
      - LINESTRING
      - MULTILINESTRING
      - POINT
      - MULTIPOINT
      - POLYGON
      - MULTIPOLYGON
      - GEOMETRYCOLLECTION
 
